---
title: "Chapter 12 - Of Monsters and Mixtures"
author: "Corrie"
date: "April 14, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Over-dispersed outcomes
For the beta-binomial model, we'll make use of the beta distribution. The beta distribution is a probability distribution over probabilities (over the interval $[0, 1]$).
```{r, warning=FALSE, message=FALSE}
library(rethinking)
pbar <- 0.5
theta <- 5
curve( dbeta2( x, pbar, theta), from=0, to=1, xlab="probability", ylab="Density")
```

There are different ways to parametrize the beta distribution:
```{r, eval=F}
dbeta2 <- function( x , prob , theta , log=FALSE ) {
    a <- prob * theta
    b <- (1-prob) * theta
    dbeta( x , shape1=a , shape2=b , log=log )
}
```

We use the beta-binomial for the `UCBadmit` data, which is over-dispersed if we ignore department (since the admission rate varied quite a lot for different departments).
Our model looks as follows:
$$\begin{align*}
A_i &\sim \text{BetaBinomial}(N_i, \bar{p}_i, \theta) \\
\text{logit}(\bar{p}_i) &= \alpha_{\text{GID}[i]}) \\
\alpha_j &\sim \text{Normal}(0, 1.5) \\
\theta &\sim \text{Exponential}(1)
\end{align*}$$
where $A$ is `admit`, $N$ is the number of applications  `applications`, and GID$[i]$ is gender id, 1 for male and 2 for female.
```{r, results='hide', warning=F, message=F}
data("UCBadmit")
d <- UCBadmit
d$gid <- ifelse(d$applicant.gender == "male", 1L, 2L)
dat <- list(A = d$admit, N = d$applications, gid=d$gid)

m12.1 <- ulam(
  alist(
    A ~ dbetabinom( N, pbar, theta),
    logit(pbar) <- a[gid],
    a[gid] ~ dnorm(0, 1.5),
    theta ~ dexp(1)
  ), data=dat, chains=4
)
```

```{r}
post <- extract.samples(m12.1 )
post$da <- post$a[,1] - post$a[,2]
precis( post, depth=2)
```

`a[1]` is the log-odds of admission for male applicants which is lower than `a[2]`, the log-odds for female applicants. This might suggest there is a diffenrece in admission rates by gender, but the posterior difference `da` is highly uncertain with probability mass both above and below zero.

We can visualize the posterior Beta-distribution:
```{r}
gid <- 2
# draw posterior mean beta distribution
curve( dbeta2(x, mean( logistic(post$a[, gid])), mean(post$theta)), from=0, to=1,
       ylab="Density", xlab="probability admit", ylim=c(0,3), lwd=2)

# draw 50 beta distributions sampled from posterior
for ( i in 1:50 ) {
  p <- logistic( post$a[i, gid ] )
  theta <- post$theta[i]
  curve( dbeta2(x, p, theta), add=T,col=col.alpha("black", 0.2))
}
mtext("distribution of female admission rates")
```

There is a tendency for admission rates below 40% but the most plausible (the mean Beta-distribution) also allows for very high admission rates above 80%. This way, the model incorporated the high variation among departments.
```{r}
postcheck(m12.1)
```

The prediction intervals are very wide.

## Negative-Binomial
Similar as with the Beta-Binomial, we can fit a Poisson model, where each observation has its own rate. The rate comes from a common gamma distribution.
```{r}
phi <- 2
lambda <- 1
curve( dgamma( x, shape=phi, rate=lambda), from=0, to=20, xlab="probability", ylab="Density")
```


```{r, results='hide', warning=F, message=F}
data(Kline)
d <- Kline
d$P <- standardize( log( d$population ) )
d$contact_id <- ifelse( d$contact == "high", 2L, 1L)

dat2 <- list(
  TT = d$total_tools,
  P = d$population,
  cid = d$contact_id
)

m12.3 <- ulam(
  alist(
    TT ~ dgampois( lambda, phi ),
    lambda <- exp( a[cid])*P^b[cid] / g,
    a[cid] ~ dnorm(1, 1),
    b[cid] ~ dexp(1),
    g ~ dexp(1),
    phi ~ dexp(1)
  ), data=dat2, chains=4, log_lik = TRUE
)
```

For comparison, our model from before:
```{r, results='hide', warning=F, message=F}
m11.11 <- ulam(
  alist(
    TT ~ dpois(lambda),
    lambda <- exp( a[cid] )*P^b[cid] / g,
    a[cid] ~ dnorm(1,1),
    b[cid] ~ dexp(1),
    g ~ dexp(1)
  ), data=dat2, chains=4, log_lik = T
)
```





```{r, fig.height=6, fig.width=11, warning=F, message=F}
k <- LOOPk(m11.11)
par(bty="l", mfrow=c(1,2))
plot( d$population, d$total_tools, xlab="population", ylab="total tools",
      col=rangi2, pch=ifelse(dat2$cid == 1, 1, 16), lwd=2,
      ylim=c(0,74), cex=1+normalize(k) )

ns <- 100
P_seq <- seq(from=0, to=12.55, length.out = ns)
pop_seq <- exp( P_seq )

lambda <- link(m12.3, data=data.frame(P = pop_seq, cid=1))
lmu <- apply(lambda, 2, mean)
lci <- apply(lambda, 2, PI )
lci2 <- ifelse(lci >= 74, 74, lci)
lines( pop_seq, ifelse(lmu>=74, NA, lmu), lty=2, lwd=1.5)
shade(lci2, pop_seq, xpd=TRUE)

lambda <- link(m12.3, data=data.frame(P = pop_seq, cid=2))
lmu <- apply(lambda, 2, mean)
lci <- apply(lambda, 2, PI )
lci2 <- ifelse(lci >= 74, 74, lci)
lines( pop_seq, ifelse(lmu>=74, NA, lmu), lty=1, lwd=1.5)
shade(lci2, pop_seq, xpd=TRUE)
text(c(9e4, 1e5), c(68, 50), labels=c("high contact", "low contact"))
mtext("gamma-Poisson model")


plot( d$population, d$total_tools, xlab="population", ylab="total tools",
      col=rangi2, pch=ifelse(dat2$cid == 1, 1, 16), lwd=2,
      ylim=c(0,74), cex=1+normalize(k) )

lambda <- link(m11.11, data=data.frame(P = pop_seq, cid=1))
lmu <- apply(lambda, 2, mean)
lci <- apply(lambda, 2, PI )
lci2 <- ifelse(lci >= 74, 74, lci)
lines( pop_seq, ifelse(lmu>=74, NA, lmu), lty=2, lwd=1.5)
shade(lci2, pop_seq, xpd=TRUE)

lambda <- link(m11.11, data=data.frame(P = pop_seq, cid=2))
lmu <- apply(lambda, 2, mean)
lci <- apply(lambda, 2, PI )
lci2 <- ifelse(lci >= 74, 74, lci)
lines( pop_seq, ifelse(lmu>=74, NA, lmu), lty=1, lwd=1.5)
shade(lci2, pop_seq, xpd=TRUE)
mtext("pure Poisson model")
```


## Zero-inflated outcomes
```{r}
prob_drink <- 0.2
rate_work <- 1

N <- 365

set.seed(365)
drink <- rbinom( N, 1, prob_drink)

y <- (1-drink)*rpois(N, rate_work)

simplehist( y, xlab="manuscripts completed", lwd=4)
zeros_drink <- sum(drink)
zeros_work <- sum(y==0 & drink == 0)
zeros_total <- sum(y==0)
lines(c(0,0), c(zeros_work, zeros_total), lwd=4, col=rangi2 )
```
