---
title: "Chapter 5 - Exercises"
author: "Corrie"
date: "June 3, 2018"
output: html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Chapter 5 - Exercises
These are my solutions to the exercises from chapter 5.

## Easy.
__5E1.__ The following linear models are multiple linear regressions:

- $\mu_i = \beta_x x_i + \beta_z z_i $
- $\mu_i = \alpha + \beta_x x_i + \beta_z z_i$
whereas the following are bivariate linear regressions:

- $\mu_i = \alpha + \beta x_i$
- $\mu_i = \alpha + \beta(x_i - z_i) $

__5E2.__ Write down a multiple regression to evaluate the claim: _Animal diversity is linearly related to latitude, but only after controlling for plant diversity._
$$
\begin{align*}
\text{animal diversity}_i &\sim \text{Normal}( \mu_i, \sigma) \\
\mu_i &= \alpha + \beta_{lat}\text{latitude}_i + \beta_{plant}\text{plant diversity}_i 
\end{align*}$$
__5E3.__ Write down a multiple regression to evaluate the claim:
_Neither amount of funding nor size of laboratory is by itself a good predictor of time to PhD degree; but together these variables are both positively associated with time to degree._

$$\begin{align*}
\text{time to PhD}_i &\sim \text{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_{f}\text{funding}_i + \beta_{l} \text{lab size}_i
\end{align*}$$
The parameters $\beta_f$ and $\beta_l$ should be both positive ($>0$).

__5E4.__ Categorical predictor with 4 levels, labeled A, B, C, and D. $A_i$, $B_i$, $C_i$, and $D_i$ are the respective indicator variables.
The following are inferentially equivalent ways to include the categorical variables in a regression:

- $\mu_i = \alpha + \beta_A A_i + \beta_B B_i + \beta_C C_i$
- $\mu_i = \alpha + \beta_B B_i + \beta_C C_i + \beta_D D_i$

## Medium.
__5M1.__

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
